Financialisation of Commodity Markets

### 2024-09-25 docs/paper.qmd
financilisation results + correlation results for crisis

Financialisation was a global phenomenon that, through the channel of index investment, simultaneously influenced the entire spectrum of liquid commodities. This sweeping impact spurred the transformation of commodity markets from a collection of idiosyncratic markets into a unified asset class with interconnectedness emerging as a crucial aspect that warrants thorough examination.   
We set out to address this issue by first examining the average of weekly returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. These findings indicate higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These findings thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-regimes.  
We further examine these correlations by year in @tbl-correlations-years to uncover any distinct breakpoints. For the full sample of assets, the average pairwise correlations increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US metals average pairwise correlations, more than doubling from 0.22 to 0.49, suggesting 2004 as the onset of financialization as noted in several studies [@baker_financialization_2021; @tang_index_2012] and our contribution is to observe that the US metals sub-sector seems to have played a major role in this.

@@

We investigate further the issue of co-movement from an asset pricing perspective using a bespoke asset pricing framework starting with a market factor mimicking portfolio constituted of the twenty-four US traded commodities considered in the study with equal weighting. This portfolio enables a refined examination of cross correlations across sectors as the numerator of the beta of the regression of the returns for an individual commodity against those for the market factor is driven by covariances while the denominator is driven by the individual commodity variance. Thus, an increase in a commodity's correlation with other commodities should lead to an increase in its beta relative to the market factor portfolio. We therefore regress the weekly individual commodity front futures returns against those for our market factor portfolio and average the resulting betas across the entire set of commodity assets considered as well as by country, sector and sub-sector for each of the four periods of interest with the results shown in @tbl-regressions-index. The average beta for the US metals sector jumps sharply to 1.12 over financialisation from 0.63 over the previous period; the increase is even more dramatic for the base metals sub-sector (GB traded assets) in a relative sense. The group had a much lower beta of 0.27 over the past period, suggesting a low level of correlation with US commodities; this figure increases to 0.85 over the financialisation period, even beyond that for the US agricultural sector of 0.80. In contrast, the corresponding figures for the energy sector show a substantial decrease, down to 1.61 during financialisation from 2.33 over the past period.  
The corresponding pricing performance displays a similar pattern as shown in @tbl-regressions-factors where for each period the corresponding $R^{2}$ values are independently averaged over the entire period as well as over periods of **contango** and **backwardation**. The average $R^{2}$ for the market factor across the entire cross-section of commodity assets considered in the study, including the GB traded base metals, increases to around 20% during financialisation compared to around 10% over the previous period with the rise driven by the metals sector at large. The sector average for the US traded assets jumps to 31% during financialisation from 7.5% over the past period while for the GB traded assets the corresponding figure increases from less than 2%, or virtually no explanatory power for the (US based) market factor to almost 12% over the financialisation period. The results for our other risk factor mimicking portfolios show a similar pattern and provide a deeper insights into co-movement dynamics in this context. All three factors---CHP, open interest and term structure---performed poorly in the first period, indicating that the commodities with extreme factor attributes (high or low CHP for example) were uncorrelated with the rest of the commodity cross-section, and the overall performance consistent across periods of **contango** and **backwardation**. The onset of financialisation leads to a substantial improvement in the performance of the CHP factor with its average $R^{2}$, albeit still small at 5.4%, almost tripling from the past to the financialisation period in contrast to the other factors that show weaker dynamics with a slight decrease and a much smaller increase for the open interest and term structure factors respectively. Here again the dramatic increase in pricing performance for the CHP factor is driven by the metals sector, particularly the precious sub-sector with the average $R^{2}$ for the group jumping to almost 26% during financialisation from 1.2% over the past period. The corresponding figures of 22% and 1.1% for the entire US metals sector stand in sharp contrast with the agricultural sector, which exhibits no notable change, and the energy sector, where pricing performance drops substantially from 9.2% down to 1.6%. The improvement in the CHP factor's pricing performance remarkably comes predominantly from the long leg (low CHP assets: Keynesian "normal backwardation") of the factor^{Results available from the authors upon request.} with the corresponding mimicking portfolio's performance itself driven by the metals, the precious sub-sector in particular. The average $R^{2}$ for the group jumps to 55% during financialisation from less than 12% over the past period while the short leg sees a far smaller increase, up to 8.9% from 3.3% over the past period. Interestingly, although the factor itself only sees a very modest increase, its short leg (low roll-yield assets: Working's backwardation) shows a sizeable rise in pricing performance, yet again driven the metals sector, with the average $R^{2}$ for the group's US traded assets jumping to 28.2% during financialisation from 6.2% over the past period. The pattern of results for both factors extends to the GB trade assets and suggest a convergence of Keneysian and Working's term structure perspective over the financialisation period.   
Taken together these results emphasise the global nature of the impact of financialisation on the metals complex and provide further evidence that global metals co-movement could be detected by US based factor models that combine the Keynesian and Working’s term structure paradigms. There seems thus to be information in the factor constructed from hedging pressure based on the Keynesian paradigm relevant to the term structure factor built on the Working paradigm and these findings therefore appear to complement the results of Kang et al. (2020) who find that hedging pressure both conveys information about liquidity provision in the short term (Working paradigm) and risk transfer in the longer term (Keynesian paradigm).

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of results. Returns fall sharply relative to the financialisation period (@tbl-stats-no-regimes) with eighteen US commodities along with all six GB traded metals showing lower mean returns. Returns for equally weighted portfolios of both sets of assets fall down to 9.4% and 5.3% from 15.9% and 21.8% respectively. Returns are mostly higher during phases of **backwardation** relative to **contango** (@tbl-stats-regimes) with fifteen US assets and five of the GB traded metals showing higher mean returns in **backwardation** and the pattern extending to equally weighted portfolios of the two sets of assets. Volatility is moderately higher for both sets with fifteen US commodities and three GB metals showing higher figures and interestingly eminently higher over phases of **contango**; out of the US complex only Sugar-#11 (IFUS) and none of the GB traded metals show higher volatility figures over **backwardation** phases. The symmetry between the dynamics of CHP and **CHP**'s explanatory power on individual commodity returns breaks in the crisis period (@tbl-time-reg-US-CHP) with both now seemingly converging; the US cross-sectional average $R^{2}$ at the individual level bouncing back up (22.5% vs 20.8%) while that at the aggregate level rises further (14.7% vs 7.1%).

// TODO Add correlation results discussion here.

>>>>

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities; this sweeping impact spurred the transformation of commodity markets from a collection of idiosyncratic markets into a unified asset class with interconnectedness emerging as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average of weekly returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicate higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-regimes.  
We further examine these correlations by year in @tbl-correlations-years to uncover any distinct breakpoints. For the full sample of assets, the average pairwise correlations increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US metals average pairwise correlations, more than doubling from 0.22 to 0.49, suggesting 2004 as the onset of financialization as noted in several studies [@baker_financialization_2021; @tang_index_2012] and our contribution is to observe that the US metals sub-sector seems to have played a major role in this.

@@

We investigate further the issue of co-movement from an asset pricing perspective using a bespoke asset pricing framework starting with a market factor mimicking portfolio constituted of the twenty-four US traded commodities considered in the study with equal weighting. This portfolio enables a refined examination of cross correlations across sectors as the numerator of the beta of the regression of the returns for an individual commodity against those for the market factor is driven by covariances while the denominator is driven by the individual commodity variance. Thus, an increase in a commodity's correlation with other commodities should lead to an increase in its beta relative to the market factor portfolio. We accordingly regress the weekly individual commodity front futures returns against those for our market factor portfolio and average the resulting betas across the entire set of commodity assets considered as well as by country, sector and sub-sector for each of the four periods of interest with the results shown in @tbl-regressions-index. The average beta for the US metals sector jumps sharply to 1.12 over financialisation from 0.63 over the previous period; the increase is even more dramatic for the base metals sub-sector (GB traded assets) in a relative sense. The group had a much lower beta of 0.27 over the past period, suggesting a low level of correlation with US commodities; this figure increases to 0.85 over the financialisation period, even beyond that for the US agricultural sector of 0.80. In contrast, the corresponding figure for the energy sector show a substantial decrease, down to 1.61 during financialisation from 2.33 over the past period.  
The corresponding pricing performance displays a similar pattern as shown in @tbl-regressions-factors where for each period the corresponding $R^{2}$ values are independently averaged over the entire period as well as over periods of **contango** and **backwardation**. The average $R^{2}$ for the market factor across the entire cross-section of commodity assets considered in the study, including the GB traded base metals, increases to around 20% during financialisation compared to around 10% over the previous period with the rise driven by the metals sector at large. The sector average for the US traded assets jumps to 31% during financialisation from 7.5% over the past period while for the GB traded assets the corresponding figure increases from less than 2%, or virtually no explanatory power for the (US based) market factor, to almost 12% over the financialisation period. The results for our other risk factor mimicking portfolios show a similar pattern and provide a deeper insights into co-movement dynamics in this context. All three factors---CHP, open interest and term structure---performed poorly in the first period, indicating that the commodities with extreme factor attributes (high or low CHP for example) were uncorrelated with the rest of the commodity cross-section, and the overall performance consistent across periods of **contango** and **backwardation**. The onset of financialisation leads to a substantial improvement in the performance of the CHP factor with its average $R^{2}$, albeit still small at 5.4%, almost tripling from the past to the financialisation period in contrast to the other factors that show weaker dynamics with a slight decrease and a much smaller increase for the open interest and term structure factors respectively. Here again the increase in pricing performance for the CHP factor is driven by the metals sector, particularly the precious sub-sector with the average $R^{2}$ for the group jumping to almost 26% during financialisation from 1.2% over the past period. The corresponding figures of 22% and 1.1% for the entire US metals sector stand in sharp contrast with the agricultural sector, which exhibits no notable change, and the energy sector, where pricing performance drops substantially from 9.2% down to 1.6%. The improvement in the CHP factor's pricing performance remarkably comes predominantly from the long leg (low CHP assets: Keynesian "normal backwardation") of the factor^{Results available from the authors upon request.} with the corresponding mimicking portfolio's performance itself driven by the metals, the precious sub-sector in particular. The average $R^{2}$ for the group jumps to 55% during financialisation from less than 12% over the past period while the short leg sees a far smaller increase, up to 8.9% from 3.3% over the past period. Interestingly, although the term structure factor itself only sees a very modest increase, its short leg (low roll-yield assets: Working's backwardation) shows a sizeable rise in pricing performance, yet again driven the metals sector, with the average $R^{2}$ for the group's US traded assets jumping to 28.2% during financialisation from 6.2% over the past period. The pattern of results for both factors extends to the GB trade assets and suggest a convergence of Keneysian and Working's term structure perspective over the financialisation period.   
Taken together these results emphasise the global nature of the impact of financialisation on the metals complex and provide further evidence that global metals co-movement could be detected by US based factor models that combine the Keynesian and Working’s term structure paradigms. There seems thus to be information in the factor constructed from hedging pressure based on the Keynesian paradigm relevant to the term structure factor built on the Working paradigm and these findings therefore appear to complement the results of @kang_tale_2020 who find that hedging pressure both conveys information about liquidity provision in the short term (Working paradigm) and risk transfer in the longer term (Keynesian paradigm).

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of results. Returns fall sharply relative to the financialisation period (@tbl-stats-no-regimes) with eighteen US commodities along with all six GB traded metals showing lower mean returns. Returns for equally weighted portfolios of both sets of assets fall down to 9.4% and 5.3% from 15.9% and 21.8% respectively. Returns are mostly higher during phases of **backwardation** relative to **contango** (@tbl-stats-regimes) with fifteen US assets and five of the GB traded metals showing higher mean returns in **backwardation** and the pattern extending to equally weighted portfolios of the two sets of assets. Volatility is moderately higher for both sets with fifteen US commodities and three GB metals showing higher figures and interestingly eminently higher over phases of **contango**; out of the US complex only Sugar-#11 (IFUS) and none of the GB traded metals show higher volatility figures over **backwardation** phases. The symmetry between the dynamics of CHP and **CHP**'s explanatory power on individual commodity returns breaks in the crisis period (@tbl-time-reg-US-CHP) with both now seemingly converging; the US cross-sectional average $R^{2}$ at the individual level bouncing back up (22.5% vs 20.8%) while that at the aggregate level rises further (14.7% vs 7.1%).   
The average pairwise correlation for the entire cross-section of commodities considered in the study (@tbl-correlations-periods) increases further into the crisis to 0.28 indicating an intensification of the higher interdependence within the full cross-section of commodities prompted by financialisation. The result is consistent with the greater interdependence between equity and commodity markets over this period documented in @silvennoinen_financialisation_2013. The figure for the US traded section of the metals sector only progresses modestly to 0.59 after peaking at 0.63 in 2006 (@tbl-correlations-years). In contrast, the GB traded assets see a sharp increase to over 0.65 in the crisis starting in 2007 up to a peak of 0.69 in 2010. The agricultural sector on the other hand shows a substantial rise in the crisis period with the sharpest increase in 2008 when the average pairwise correlation for the sector jumps to 0.30. Meanwhile, the energy sector also shows a notable progression starting in 2007 and peaking at 0.63 in 2008 before levelling back down over the rest of the period. The onset of the financial crisis seems therefore to have had an across the board impact on the interdependence of commodities in contrast with financilisation, much more closely related to the metals sector.
